Verification of concurrent design patterns with data

### 2019-04-26 2018-Arch-pNets/Coordination Camera ready/arch-pNets.tex
arch-pNets paper: Final camera-ready version to be submitted

@@

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$ and let $\sigma$ range over valuations.
Valuations extend canonically to expressions, denoted $\val{}{}(e)$. We define:
%\noteLH{its the following really necessary}
%\addSB{Since we consider variables to be sorted, valuations
%  applied to expressions are interpreted as partial functions
%  $\exprs{V} \rightharpoonup \data$.}
%For a valuation $\val{}{}$ and an assignment $(x_i := e_i)^{i \in I}$, we define:
\vspace{-1.5ex}
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}\vspace{-1ex}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e$ is  \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} $ is
 monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We  denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%
%****************************************************************

@@

%\paragraph{Definitions.}
%\label{section:pNets}
A pLTS is a labelled transition system with variables occurring inside states,  actions, guards, and
assignments. 
%Without loss of generality and to simplify the formalisation, we suppose 
%here that 
%variables are local to each 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
%Note that we make no assumption on the finiteness of the set of states, nor
%on the finite branching of the transition relation.
Variables of each state are pairwise disjoint.
\newCoord{The parameters of actions that appear on pLTS transitions are either input variables, bound when the action occurs and accepting any value (of the right type), or expressions using the value of other variables.}
We  define the set of parameterised actions a pLTS can use ($a$
ranges over action labels):
% Action terms are  of the form 
$\alpha=a(?x_i^{i \in I}, 
e_j^{j \in J})$, where $?x_i^{i \in I}$ are input variables, $e_j^{j \in J}$ 
are expressions.

@@

\newCoord{The red dotted rule expresses the implication stating that if the holes perform the designated actions and the condition $g$ is verified, then the variables are modified and the state changes. This implication however  uses a simple logic with the 
 expressive power  given by the predicate algebra (it must include logical 
 propositions and equality). 
 Proposition and inference rules of the paper use a standard logic, while predicates inside the open transitions should use a more restricted logic, typically a logic that could  be handled mechanically and expressed by terms that can be encoded in a simple syntax.
Open transitions express in a symbolic way, transitions  that are not only parametrised with variables but also actions of not yet known processes.}

@@

\begin{example}[An open-transition]
  \label{ex:TimerOT3}
We now briefly illustrate an open transition:
\begin{small}\[ \openrule
  {\{B\mapsto fail(b)\},
    t<Max \land b=true ,
    \{t' := t+1\} }
  {\ostate{11} \OTarrow{\alpha(t)} \ostate{14}}\vspace{-1.3ex}

@@

It requires the hole at label $B$ to fire a $fail(b)$ action, with the condition $b=true$ and the state variable $t<Max$. In this case, the global pNet goes from state $\ostate{11}$ that has an internal variable $t$ to state $\ostate{14}$ where the state variable $t'$ has the value $t+1$. This emits a global action $\alpha(t)$. \newCoord{Figure~\ref{fig:ArchFailure:OA} illustrates a complete open automaton.}

@@

\subsubsection{Components and composition}
\label{secn:components}
\begin{definition}[Component]
  \label{defn:component}
  A component is a tuple $\tuple{Q, q^0, V, \val{0}{}, P,
  \export, \goesto{}}$ where\vspace{-2ex}
  \begin{itemize}
  \item $Q$ is a set of \emph{states}, with $q^0 \in Q$ the
    \emph{initial state}, 
  \item $V$ is a set of \emph{component variables},
  \item $\val{0}{} : V \rightarrow \data$ is an \emph{initial
    valuation} of the component variables, 
  \item $P$ is a set of \emph{ports};
    $\export : P \rightarrow 2^V$ is
    the \emph{set of variables exported by each port},
  \item $\goesto{}\, \subseteq
    Q \times (2^P\setminus \{\emptyset\}) \times \guards{V} \times \updates{V} \times Q$
%
    is a \emph{transition relation}, with transitions
    labelled by \emph{interactions}, \ie triples consisting of a non-empty set of ports,
    %% $\emptyset \neq a \subseteq P$,
    a monotonic Boolean \emph{guard}
    %% $g \in \guards{V}$ 
    and a monotonic \emph{update assignment}.
    %% \hl{$u \in \updates{V}$}.
  \end{itemize}
%
  We call the triple $(V,P, \export)$ the \emph{interface} of the
  component.\,\footnote{Only exported variables, belonging to a
  $\export(p)$, appear in the component interface  (see
  \defn{im}).  We omit here this separation between internal and exported variables.
  }
%
  Notations $q \goesto{a, g, u} q'$ and $q \goesto{a, g, u}$ are as usual;
  for a component $B$, we denote $Q_B$, $q^0_B$, $V_B$, $\val{0}{B}$, $P_B$,
  and $\export[B]$ the corresponding constituents of $B$.  We will
  skip the index on the transition relations $\goesto{}$, since it
  is always clear from the context.
\end{definition}

@@

\begin{definition}[Component semantics]
  \label{defn:comp:semantics}
  The \emph{open semantics} of a component $B = \tuple{Q, q^0, V,
  \val{0}{}, P, \export, \goesto{}}$ is %% given by
  the LTS denoted $\semopen{B} = \tuple{S,
  s^0, \goesto{}}$, where $S = Q \times \valuations{V}$, $s^0 =
  \tuple{q^0, \val{0}{}}$ and $\goesto{}$ is the minimal transition
  relation satisfying the rule
  %
  \begin{equation}
    \label{eq:comp:semantics}
    \infer{
      q \goesto{a, g, u} q'
      \and
      \val{}{} \models g
      \and
      \val[\primeit]{}{} = \tilde{\val{}{}}[u]
      \and
      \valdiff{\val{}{}}{\tilde{\val{}{}}} \subseteq \export(a)
    }{
      (q, \val{}{}) \goesto{a,\tilde{\val{}{}}} (q', \val[\primeit]{}{})
    }
    \,.
  \end{equation}
%
  The \emph{closed semantics} of $B$ is given by the LTS denoted
  $\semclosed{B}$, comprising only those transitions of $\semopen{B}$,
  where $\tilde{\val{}{}} = \val{}{}$, which we denote
  $(q, \val{}{}) \goesto{a} (q', \val[\primeit]{}{})$.
\end{definition}

@@

\begin{definition}[Architecture]
  \label{defn:arch}
  An \emph{architecture} is a tuple $A = \tuple{\cC, V_A, P_A,
    \export[A], \Gamma}$, where\vspace{-1.5ex}
  \begin{itemize}
  \item $P_A$ and $V_A$ are sets of ports and variables, respectively;
\item
  $\cC$ is a finite set of components, which we call \emph{coordinators},
  %% with pairwise disjoint sets of ports and variables,
  such that
  $\bigcup_{C \in \cC} P_C \subseteq P_A$ and
  $\bigcup_{C \in \cC} V_C \subseteq V_A$;
%
  ports in $P_A \setminus \bigcup_{C \in \cC} P_C$, which do not
  belong to any of the coordinators are called
  \emph{dangling};

@@

\begin{definition}[Composition of architectures]
  \label{defn:arch:composition}
  Let $A_i = \tuple{\cC_i, V_{A_i}, P_{A_i}, \export[A_i], \Gamma_i}$, for $i = 1,2$,
  be two architectures.  The \emph{composition} of $A_1$ and
  $A_2$ is the architecture $A_1 \arcomp A_2 = \tuple{\cC_1 \cup \cC_2,
  V_{A_1} \cup V_{A_2}, P_{A_1} \cup P_{A_2}, \export[A_1] \cup \export[A_2], \Gamma}$, where\vspace{-1ex}
%
  \begin{equation}
    \label{eq:arch:composition}
    \Gamma = \bsetdef{
      (a, g^1 \land g^2, u^1 \expmix u^2) 
    }{
      (a \cap P_{A_i}, g^i, u^i) \in \Gamma_i,
      \text{ for } i = 1,2
    }
    \,.
  \end{equation}
\end{definition}

@@

\section{Related work}
\label{secn:related}
%% Aside from previous work on pNets and BIP architectures mentioned in \secn{introduction}. We  review here the closest related works outside the scope of pNets and BIP architectures, with a particular focus on compositional and parameterized verification.

@@

%% infinite/parametrized model checking? == ERIC (listes ci-dessous)  + Simon

>>>>

\newcommand{\SVTuple}[2][\ensuremath{,\allowbreak}]{%
  \setTupleDelims[\langle]{\rangle}%
  \tuplebase[#1]{#2}%
}

@@

For a set of variables $V \subseteq \variables$, we denote
$\valuations{V} \bydef{=} \{\val{}{}: V \rightarrow \data\}$
the set of \emph{valuations} of the variables in $V$ and let $\sigma$ range over valuations.
Valuations extend canonically to expressions, denoted $\val{}{}(e)$. We define:
%\noteLH{its the following really necessary}
%\addSB{Since we consider variables to be sorted, valuations
%  applied to expressions are interpreted as partial functions
%  $\exprs{V} \rightharpoonup \data$.}
%For a valuation $\val{}{}$ and an assignment $(x_i := e_i)^{i \in I}$, we define:
%% \vspace{-1.5ex}
\[
\val{}{}\bigl[(x_i := e_i)^{i \in I}\bigr](x) \bydef{=}
\begin{cases}
  \val{}{}(x), & \text{if } x \not\in x_i^{i \in I}\,,\\
  \val{}{}(e_i), & \text{if } x = x_i, \text{for some } i \in I\,.
\end{cases}%% \vspace{-1ex}
\]
For two valuations $\val{1}{}, \val{2}{} : V \rightarrow
\data$, we denote $\valdiff{\val{1}{}}{\val{2}{}} \bydef{=}
\bsetdef{x \in V}{\val{1}{}(x) \neq \val{2}{}(x)}$ the set of
variables that are assigned different values by the two
valuations.  As usual, we write $\val{1}{} \order \val{2}{}$
iff $\val{1}{}(x) \order \val{2}{}(x)$, for all $x \in V$.  An
expression $e$ is  \emph{monotonic} if, for
any two valuations $\val{1}{}, \val{2}{}$,
$\val{1}{} \order \val{2}{}$ implies $\val{1}{}(e) \order
\val{2}{}(e)$.
%
Similarly, an assignment $(x_i := e_i)^{i \in I} $ is
 monotonic if all expressions $e_i^{i \in I}$ are monotonic.
We  denote $\guards{V} \subset \boolexprs{V}$, $\monexprs{V}
\subset \exprs{V}$ and $\updates{V} \subset \assigns{V}$ the sets of
monotonic Boolean and generic expressions and assignments,
respectively.
%
%****************************************************************

@@

%\paragraph{Definitions.}
%\label{section:pNets}
A pLTS is a labelled transition system with variables occurring inside states,  actions, guards, and
assignments. 
%Without loss of generality and to simplify the formalisation, we suppose 
%here that 
%variables are local to each 
%state: each state has its set of variables disjoint from the others. Transmitting 
%variable values from one state to the other can be done by explicit assignment. 
%Similarly, to simplify the management of variables and without loss of expressivity, we 
%suppose that transitions looping to the same state does not do assignments.
%Note that we make no assumption on the finiteness of the set of states, nor
%on the finite branching of the transition relation.
Variables of each state are pairwise disjoint.
\newCoord{Each transition label of a pLTS consists of a parameterised action, a guard and an update assignment.  The parameters of actions are either input variables or expressions.  Input variables are bound when the action occurs; they accept any value (of the correct type), thus providing a to input data from the environment.  Expressions are computed from the values of other variables.  They allow providing aggregated values to the environment, without exposing all the underlying variables.}
We  define the set of parameterised actions a pLTS can use ($a$
ranges over action labels):
% Action terms are  of the form 
$\alpha=a(?x_i^{i \in I}, 
e_j^{j \in J})$, where $?x_i^{i \in I}$ are input variables, $e_j^{j \in J}$ 
are expressions.

@@

\newCoord{The red dotted rule expresses the implication stating that if the holes perform the designated actions and the condition $g$ is verified, then the variables are modified and the state changes. This implication however  uses a simple logic with the 
 expressive power  given by the predicate algebra (it must include logical 
 propositions and equality). 
 Proposition and inference rules of the paper use a standard logic, while predicates inside the open transitions should use a more restricted logic, typically a logic that could  be handled mechanically and expressed by terms that can be encoded in a simple syntax.
Open transitions express in a symbolic way, transitions  that are not only parameterised with variables but also actions of not yet known processes.}

@@

\begin{example}[An open transition]
  \label{ex:TimerOT5}
  \newCoord{The open transition
  \begin{small}
    \[
    \openrule{
      \{E\mapsto ask\},
      t \in z,
      \{t:=t+1\}
    }{
      \ostate{11} \OTarrow{ask} \ostate{11}
    }

@@

%% \begin{small}\[ \openrule
%%   {\{B\mapsto \mathit{fail}(b)\},
%%     t<Max \land b=true ,
%%     \{t := t+1\} }
%%   {\ostate{11} \OTarrow{\alpha(t)} \ostate{14}}%% \vspace{-1.3ex}
%%   \]
%% \end{small}
emits a global action $ask$ labelling the synchronisation vector $\SVTuple{\mathit{timeout}_T, \mathit{timeout}_C, -, \mathit{ask}} \rightarrow \mathit{ask}$.  It requires the hole at label $E$ to fire an $\mathit{ask}$ action, with the condition $t \in z$. In this case, the global pNet loops on the state $\ostate{11}$ that has internal variables $t$ and $z$ local to the pLTS $T$ (hence not appearing in the synchronisation vector).  The variable $t$ is updated to the new value $t+1$. \Fig{FailureTimer:pNet} shows the complete pNet, whereas \fig{ArchFailure:OA} shows a complete open automaton.}

@@

\subsubsection{Components and composition}
\label{secn:components}
\begin{definition}[Component]
  \label{defn:component}
  A component is a tuple $\tuple{Q, q^0, V, \val{0}{}, P,
  \export, \goesto{}}$ where%% \vspace{-2ex}
  \begin{itemize}
  \item $Q$ is a set of \emph{states}, with $q^0 \in Q$ the
    \emph{initial state}, 
  \item $V$ is a set of \emph{component variables},
  \item $\val{0}{} : V \rightarrow \data$ is an \emph{initial
    valuation} of the component variables, 
  \item $P$ is a set of \emph{ports};
    $\export : P \rightarrow 2^V$ is
    the \emph{set of variables exported by each port},
  \item $\goesto{}\, \subseteq
    Q \times (2^P\setminus \{\emptyset\}) \times \guards{V} \times \updates{V} \times Q$
%
    is a \emph{transition relation}, with transitions
    labelled by \emph{interactions}, \ie triples consisting of a non-empty set of ports,
    %% $\emptyset \neq a \subseteq P$,
    a monotonic Boolean \emph{guard}
    %% $g \in \guards{V}$ 
    and a monotonic \emph{update assignment}.
    %% \hl{$u \in \updates{V}$}.
  \end{itemize}
%
  We call the triple $(V,P, \export)$ the \emph{interface} of the
  component.\,\footnote{Only exported variables, belonging to a
  $\export(p)$, appear in the component interface  (see
  \defn{im}).  We omit here this separation between internal and exported variables.
  }
%
  Notations $q \goesto{a, g, u} q'$ and $q \goesto{a, g, u}$ are as usual;
  for a component $B$, we denote $Q_B$, $q^0_B$, $V_B$, $\val{0}{B}$, $P_B$,
  and $\export[B]$ the corresponding constituents of $B$.  We will
  skip the index on the transition relations $\goesto{}$, since it
  is always clear from the context.
\end{definition}

@@

\begin{definition}[Component semantics]
  \label{defn:comp:semantics}
  The \emph{open semantics} of a component $B = \tuple{Q, q^0, V,
  \val{0}{}, P, \export, \goesto{}}$ is %% given by
  the LTS denoted $\semopen{B} = \tuple{S,
  s^0, \goesto{}}$, where $S = Q \times \valuations{V}$, $s^0 =
  \tuple{q^0, \val{0}{}}$ and $\goesto{}$ is the minimal transition
  relation satisfying the rule
  %
  \begin{equation}
    \label{eq:comp:semantics}
    \infer{
      q \goesto{a, g, u} q'
      \and
      \val{}{} \models g
      \and
      \val[\primeit]{}{} = \tilde{\val{}{}}[u]
      \and
      \valdiff{\val{}{}}{\tilde{\val{}{}}} \subseteq \export(a)
    }{
      (q, \val{}{}) \goesto{a,\tilde{\val{}{}}} (q', \val[\primeit]{}{})
    }
    \,.
  \end{equation}
%
  \newCoord{%
    The \emph{closed semantics} of $B$ is given by the LTS denoted
  $\semclosed{B}$, comprising only those transitions of $\semopen{B}$,
  where $\tilde{\val{}{}} = \val{}{}$.}%% , which we denote
  %% $(q, \val{}{}) \goesto{a} (q', \val[\primeit]{}{})$.}
\end{definition}

@@

\begin{definition}[Architecture]
  \label{defn:arch}
  An \emph{architecture} is a tuple $A = \tuple{\cC, V_A, P_A,
    \export[A], \Gamma}$, where%% \vspace{-1.5ex}
  \begin{itemize}
  \item $P_A$ and $V_A$ are sets of ports and variables, respectively;
\item
  $\cC$ is a finite set of components, which we call \emph{coordinators},
  %% with pairwise disjoint sets of ports and variables,
  such that
  $\bigcup_{C \in \cC} P_C \subseteq P_A$ and
  $\bigcup_{C \in \cC} V_C \subseteq V_A$;
%
  ports in $P_A \setminus \bigcup_{C \in \cC} P_C$, which do not
  belong to any of the coordinators are called
  \emph{dangling};

@@

\begin{definition}[Composition of architectures]
  \label{defn:arch:composition}
  Let $A_i = \tuple{\cC_i, V_{A_i}, P_{A_i}, \export[A_i], \Gamma_i}$, for $i = 1,2$,
  be two architectures.  The \emph{composition} of $A_1$ and
  $A_2$ is the architecture $A_1 \arcomp A_2 = \tuple{\cC_1 \cup \cC_2,
  V_{A_1} \cup V_{A_2}, P_{A_1} \cup P_{A_2}, \export[A_1] \cup \export[A_2], \Gamma}$, where%% \vspace{-1ex}
%
  \begin{equation}
    \label{eq:arch:composition}
    \Gamma = \bsetdef{
      (a, g^1 \land g^2, u^1 \expmix u^2) 
    }{
      (a \cap P_{A_i}, g^i, u^i) \in \Gamma_i,
      \text{ for } i = 1,2
    }
    \,.
  \end{equation}
\end{definition}

@@

\section{Related work}
\label{secn:related}
%% Aside from previous work on pNets and BIP architectures mentioned in \secn{introduction}. We  review here the closest related works outside the scope of pNets and BIP architectures, with a particular focus on compositional and parameterised verification.

@@

%% infinite/parameterised model checking? == ERIC (listes ci-dessous)  + Simon
